---
title: "hw4_discussion"
output: html_document
date: "2023-02-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
student <- read.csv("~/Downloads/student (1).csv")
head(student)
```



(a) Using R, find the lower bound for the 95% confidence
interval for the average students height.

```{r}
heightCI = t.test(student$height, conf.level = .95)
heightCI$conf.int[1] #1 for lower bound
```


(b) Using R, find the upper bound for the 95% confidence interval for the average students height.

```{r}
heightCI = t.test(student$height, conf.level = .95)
heightCI$conf.int[2] #2 for upper bound
```

(c) What is the highest average height you expect a student to have, based on the confidence interval from
a/b?
It is 68.5639.

(d) Using R, find the lower bound for the 99% confidence
interval for the average students pulse.
```{r}
pulseCI = t.test(student$pulse, conf.level = .99)
pulseCI$conf.int[1] #1 for lower bound
```

(e) Using R, find the upper bound for the 99% confidence interval for the average students pulse.
```{r}
pulseCI = t.test(student$pulse, conf.level = .99)
pulseCI$conf.int[2] #1 for lower bound
```

(f) What is the lowest average pulse you expect a student to have, based on the confidence interval from
d/e?

The lowest average pulse we expect is 71.5894. 

# Problem II 

```{r}
Radish <- read.csv("~/Downloads/Radish.csv")
head(Radish)
```

(a) Find the lower bound for the 90% confidence interval
for the difference between the two groups.
```{r}
radishCI = t.test(Radish$Height ~ Radish$Treatment, conf.level = .90)
radishCI
```

```{r}
radishCI$conf.int[1]
```


(b) Find the upper bound for the 90% confidence interval for the difference between the two groups.

```{r}
radishCI$conf.int[2]
```


(c) The fertilizer manufacturer claims that by using
their fertilizer, plants will grow more on average than
if you use no fertilizer. Does your confidence interval
support this claim?

No, because 0 is not in the confidence interval meaning that the true mean difference of control - fertilizer seems to be greater than 0. 

Pretty much the fertilizer makes it worse.

(d) What is the largest difference in average growth you
could expect with 90% confidence, based on your
confidence interval?

0.8507992


