---
title: "Template for Project 1 Data Analysis Report (replace it with your own title)"
author: "(Your name)"
date: "(Date on which the report is written)"
output:
  html_document:
    df_print: paged
    number_sections: yes
---
```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.pos = 'H')
```

***

```{r}
#install.packages("qwraps2", repo = "http://cran.rstudio.com")
library(qwraps2)
packageVersion("qwraps2")
library(data.table)
```



Note: 


- You may use other templates or formats that you are comfortable with. 

- **Remove these remarks in your submission.**

***

# Abstract 

<span style='color:blue'> 
A (very) short summary of the report. As an example, you can simply write one sentence for each section in the report. 
</span>




# Introduction

<span style='color:blue'> In this section, state the questions of interest, motivation of this analysis, and potential impact of your results. You can simply rephrase the Project Description for minimal efforts. You can also cite published papers or credible articles on the Internet. For instance, you may find  [this brief](https://eric.ed.gov/?id=ED540485) very relevant. More can be found by searching the key words "class size", "education", "performance." See, among others,[here](https://owl.purdue.edu/owl/research_and_citation/apa_style/apa_formatting_and_style_guide/in_text_citations_the_basics.html) for proper citation formats.   </span>
 
# Background 

<span style='color:blue'> In this section, explain the source of data, target population, sampling mechanism, and variables in this data set. You can briefly review  existing research or known results, which will help you in the analysis. You can find the data set from many sources, e.g.,  [the AER package](https://www.rdocumentation.org/packages/AER/versions/1.2-9/topics/STAR), [Harvard dataverse](https://dataverse.harvard.edu/dataset.xhtml?persistentId=hdl:1902.1/10766). Both links provides information about this dataset. There are different challenges using data from these two sources (see descriptive analysis).

[The brief](https://eric.ed.gov/?id=ED540485) mentions in previous section is also a good reference to read. You can find more by searching the key word "Project STAR" in, e.g.,  [Google scholar](https://scholar.google.com/scholar?hl=en&as_sdt=0%2C5&q=Project+STAR&btnG=).
</span> 

# Descriptive analysis 


Load the data 




```{r}
df <- read.delim("~/STA-207/STAR_Students.tab", header=TRUE)
```

```{r}
head(df)
```




We should drop all the rows and columns that are irrelevant like the non-first grade scores. 


```{r}
#drop the N/A values rows in g1tchid
df <- subset(df, g1tchid > 0)
```


```{r}
names(df)
```

```{r}
#won't overthink and just focus on what we need and then go back and reevaluate 
df <- df[c("stdntid", "g1tchid", "g1schid", "g1classtype",  "g1classsize", "g1tmathss")]
df
```

```{r}
#drop any with N/A values
df <- na.omit(df)
df
```
 

```{r}
#these identify the classes 
unique(df$g1tchid)
```

```{r}
#all the students in the 23462807 class, notice one probably had an N/A and was dropped
df[df$g1tchid == "23462807",]
```

Based on preliminary reading here https://eric.ed.gov/?id=ED540485
there were are 3 class types:  ‘Small’ classes (about 15-17
students), ‘Regular’ classes (about 22-25 students), and ‘Regular with a full-time Aide’ classes
(about 22-25 students)


We will summarize the data with by splitting it into the three groups and then using quantiles since they robust when dealing with outliers 

We can also have a section showing how many teachers and students there are in our data set 

Teacher could have multiple classes 

```{r}
our_summary_small <-
  list("math scaled scores in 1st grade" =
         list(min = ~ min(g1tmathss),
              q1 = ~ quantile(g1tmathss, 0.25),
              median = ~ quantile(g1tmathss, 0.25),
              q3 = ~ quantile(g1tmathss, 0.25),
              max = ~ max(g1tmathss)))
```


Group the students by the teacher and give the 5 number summary for the class

```{r}
grouped_by_table <- summary_table(df, our_summary_small, by = "g1tchid")
t(grouped_by_table)
```

Outcome vs class type 

We will say the outcome is their score 

```{r}
boxplot(g1tmathss ~ g1classtype,data = df, title = "class type vs outcome")
```

We notice that the small class type seems to have better scores on the test. 


```{r}
our_summary_small <-
  list("math scaled scores in 1st grade" =
         list(min = ~ min(g1tmathss),
              q1 = ~ quantile(g1tmathss, 0.25),
              median = ~ quantile(g1tmathss, 0.25),
              q3 = ~ quantile(g1tmathss, 0.25),
              max = ~ max(g1tmathss)))
```


Group the students by the teacher and give the 5 number summary for the class

```{r}
grouped_by_table <- summary_table(df, our_summary_small, by = "g1schid")
t(grouped_by_table)
```






<span style='color:blue'> 
Select the variables you find relevant based on your understanding in the Background section. Summarize univariate descriptive statistics for the selected variables (mean, standard deviations, missing values,  quantiles, etc.). You can create the table using functions in base `R`, or use packages (see, e.g., [this note](https://cran.r-project.org/web/packages/qwraps2/vignettes/summary-statistics.html)). </span>


<span style='color:blue'> 
From the data set, we can easily notice that various number of students are assigned to each teacher. In order to obtain one summary measure with teacher as the unit, we need to aggregate students' performance (their math scores in 1st grade). </span>

- Choose the summary measure to be used. Your options are the mean, median, quantiles, etc. Be sure to justify your choice. 
- Calculate this summary measure for each teacher. You may find the  `summarise()` function helpful ([link](https://dplyr.tidyverse.org/reference/summarise.html)). 


<span style='color:blue'>
Depending on whether you use the data from `AER` or from the Harvard dataverse. You may run into different issues in preprocessing. 
</span>

- `AER`: There are no variables that present teacher ID or class ID. However, it is possible to uniquely identify teachers/classes based on variables `experience1`,  `tethnicity1`, `schoolid1`, and `star1`. 

- Harvard dataverse: You need to read the description to find and download the data set. The data set is in the `.sav` format with 379 variables. However, you can easily identify teacher/class in 1st grade using the variable `g1tchid`. A copy of the data set is available [here](https://www.dropbox.com/s/kbehsfec44zvzod/STAR_Students.sav?dl=0) in case the server crashes.



<span style='color:blue'> 
Multivariate descriptive statistics for the outcome (the chosen summary measure for each teacher) with key variables (e.g., class types, school IDs). </span>

- Outcome v.s. class types: You can draw boxplots using `ggplot2` ([link](https://ggplot2.tidyverse.org/reference/geom_boxplot.html)). 
- Outcome v.s. school IDs: you may want to report selected summary statistics, since there are many schools and a handful of teachers/classes per school. 
 

# Inferential analysis 


<span style="color:blue">
We can define a two-way ANOVA model as follows 
$Y_{ijk} = \mu_{..} + \alpha_{i} + \beta_{j} + \epsilon_{ijk}$, where the index $i$ represents the class type: small ($i=1$), regular ($i=2$), regular with aide ($i=3$), and the index $j$ represents the school indicator. You need to explain the rest of the parameters, state constraints on the parameters, and justify the choice of model (e.g., why no interaction terms).  
</span>


<span style="color:blue">
State the assumptions on your proposed model.
</span>


<span style="color:blue">
Fit the model on the Project STAR data and report the fitted results with some attention on how/whether to report the estimated coefficients for school IDs. 
</span>


<span style="color:blue">
State the hypotheses to test. Please be sure specify the significance level and interpret your test result.  Explain any additional assumptions involved in this test. 
</span>


<span style='color:blue'>
For the secondary question of interest, one option is the Tukey's range test ( [link](https://en.wikipedia.org/wiki/Tukey%27s_range_test)). You can employ other testing procedure as well.
Again, specify the significance level, interpret your test result, and explain any additional assumptions involved in this test.  
</span> 

# Sensitivity analysis 


- <span style='color:blue'> Examine the residual plot of the fitted model.  You need to explain your findings in these plots (e.g., whether assumptions are plausible). </span>

- <span style='color:blue'>  You can find tests for some assumptions by searching the key words "test" and the corresponding assumptions. 

- <span style='color:blue'>  For alternative methods, you can explore 
  - other summary measures (say, median instead of mean)
  - or nonparametric approach and check if your answers to the questions of interest change. 
</span>


# Discussion 

<span style='color:blue'> 
Conclude your analysis in this section. You can touch on the following topics. 
</span> 

- A brief recap of this project. 
- Findings in the inferential analysis interpreted in the context of Project STAR. 
- Suggestions for future research and/or policy making given your findings. 
- Caveats of the current analysis.

# Acknowledgement {-}

<span style='color:blue'>
By default, it is assumed that you have discussed this project with your teammates and instructors. List any other people that you have discussed this project with. 
</span>

# Reference {-}

<span style='color:blue'>
List any references you cited in the report. See [here](https://owl.purdue.edu/owl/research_and_citation/apa_style/apa_formatting_and_style_guide/in_text_citations_the_basics.html) for the APA format, as an example: 
</span> 

Imbens, G., & Rubin, D. (2015). Stratified Randomized Experiments. In Causal Inference for Statistics, Social, and Biomedical Sciences: An Introduction (pp. 187-218). Cambridge: Cambridge University Press. doi:10.1017/CBO9781139025751.010

# Session info {-}

<span style='color:blue'>
Report information of your `R` session for reproducibility. 
</span> 


```{r}
sessionInfo()
```